---
title: "Introduction to R"
output:
  html_document:
    theme: cerulean
---

<link rel="stylesheet" type="text/css" href="style.css">

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```

<!-- This file was adapted by Aleeza Gerstein from the attributions listed below. It is licensed under a Creative Commons Attribution 4.0 International License. -->

---

<div class="alert alert-info">
<strong>Learning Objectives</strong>

*  Define the following terms as they relate to R: object, assign, call, function, arguments, options.   
*  Assign values to objects in R.   
*  Learn how to name objects   
*  Use comments to inform script.   
*  Solve simple arithmetic operations in R.   
*  Call functions and use arguments to change their default options.   
*  Inspect the content of vectors and manipulate their content.   
*  Subset and extract values from vectors.   
*  Analyze vectors with missing data.   
</div>

---


# Creating objects in R

You can get output from R simply by typing math in the console:

```{r}
3 + 5
12 / 7
```

However, to do useful and interesting things, we need to assign values to objects. To create an object, we need to give it a name followed by the assignment operator <-, and the value we want to give it:

```{r}
weight_kg <- 55
```

<- is the *assignment* operator. It assigns values on the right to objects on the left. So, after executing  `x <- 3`, the value of `x` is `3`. The arrow can be read as 3 **goes into** x. For historical reasons, you can also use `=` for assignments, but not in every context. Because of the slight differences in syntax, it is best practice to always use `<`- for assignments.

Objects can be given any name such as `x`, `current_temperature`, or `subject_id`. You want your object names to be explicit and not too long. They cannot start with a number (`2x` is not valid, but `x2` is). R is case sensitive (e.g., `weight_kg` is different from `Weight_kg`). There are some names that cannot be used because they are the names of fundamental functions in R (e.g., `if`, `else`, `for`, see ["here"](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html) for a complete list). In general, even if it’s allowed, it’s best to not use other function names (e.g., `c`, `mean`, `data`,  `weights`). If in doubt, check the help to see if the name is already in use. It’s also best to avoid dots (`.`) within an object name as in `my.dataset`. There are many functions in R with dots in their names for historical reasons, but because dots have a special meaning in R (for methods) and other programming languages, it’s best to avoid them. It is also recommended to use nouns for object names, and verbs for function names. 

It’s important to be consistent in the styling of your code (where you put spaces, how you name objects, etc.). Using a consistent coding style makes your code clearer to read for your future self and your collaborators. In R, three popular style guides are ["Google’s"](https://google.github.io/styleguide/Rguide.xml), ["Jean Fan’s"]("http://jef.works/R-style-guide/") and the ["tidyverse’s"](https://style.tidyverse.org/). 
This may seem overwhelming at first. My suggestion is to focus on what makes sense to you, and that you are using naming conventions that will help you to remember what is stored in each object. Try and be consistent with spacing and capitalization (or not) --- your future self will thank you.

# Objects vs. variables

What are known as `objects` in `R` are known as `variables` in many other programming languages. In this lesson, the two words are used synonymously. For more information see https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects

When assigning a value to an object, R does not print anything. You can force R to print the value by using parentheses or by typing the object name.

```{r}
weight_kg <- 55   #  doesn't print anything
(weight_kg <- 55) #  will print the value of `weight_kg`
weight_kg         #  and so does typing the name of the object
```

Now that R has `weight_kg` in memory, we can do arithmetic with it. For instance, we may want to convert this weight into pounds (weight in pounds is 2.2 times the weight in kg):

```{r}
2.2 * weight_kg
```

We can also change an object’s value by assigning it a new one:

```{r}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one object does not change the values of other objects For example, let’s store the animal’s weight in pounds in a new object, `weight_lb`:

```{r}
weight_lb <- 2.2 * weight_kg
```

and then change weight_kg to 100.

```{r}
weight_kg <- 100
```

What do you think is the current content of the object `weight_lb`? 126.5 or 220?

### Comments

The comment character in `R` is `#`, anything to the right of a `#` in a script will be ignored by R. It is useful to leave notes, and explanations in your scripts. RStudio makes it easy to comment or uncomment a paragraph: after selecting the lines you want to comment, press at the same time on your keyboard `Ctrl + Shift + C`. If you only want to comment out one line, you can put the cursor at any location of that line (i.e. no need to select the whole line), then press `Ctrl + Shift + C`.

<div class="box">
### EXERCISE
What are the values stored in each object in the following (do this in your head)?

```{r}
mass <- 50              # mass?
age  <- 122             # age?
mass <- mass * 2.0      # mass?
age  <- age - 22        # age?
mass_index <- mass/age  # mass_index?
```
</div>

# Functions and their arguments

Functions are “canned scripts” that automate more complicated sets of commands including operations assignments, etc. Many functions are predefined, or can be made available by importing R packages (more on that later). 


A function usually takes one or more inputs called arguments. Functions often (but not always) return a value. A typical example would be the function `sqrt()`. The input (the argument) must be a number, and the return value (in fact, the output) is the square root of that number. 

You can think of a function call generally as

```
do_this(to_that)
do_that(to_this, to_that, with_those)
```

Executing a function (‘running it’) is called calling the function. An couple of examples of function calls are:

```{r}
print("hello world")
```


```{r}
a <- 9
b <- sqrt(a)
b
```

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()` function calculates the square root, and returns the value which is then assigned to the object `b`. This function is very simple, because it takes just one argument.

The return ‘value’ of a function need not be numerical (like that of `sqrt()`), and it also does not need to be a single item: it can be a set of things, or even a dataset. We’ll see that when we read data files into R.

Arguments can be anything, not only numbers or filenames, but also other objects. Exactly what each argument means differs per function, and must be looked up in the documentation (see below). Some functions take arguments which may either be specified by the user, or, if left out, take on a default value: these are called options. Options are typically used to alter the way the function operates, such as whether it ignores ‘bad values’, or what symbol to use in a plot. However, if you want something specific, you can specify a value of your choice which will be used instead of the default.

Let’s try a function that can take multiple arguments: `round()`.

```{r}
round(3.14159)
```

Here, we’ve called `round()` with just one argument, `3.14159`, and it has returned the value 3. That’s because the default is to round to the nearest whole number. If we want more digits we can see how to do that by getting information about the round function. We can use `args(round)` to find what arguments it takes, or look at the help for this function using `?round`.

```{r}
args(round)
```

```{r}
?round
```

We see that if we want a different number of digits, we can type `digits = 2` or however many we want.

```{r}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined you don’t have to name them:

```{r}
round(3.14159, 2)
```

And if you do name the arguments, you can switch their order:

```{r}
round(digits = 2, x = 3.14159)
```

It’s good practice to put the non-optional arguments (like the number you’re rounding) first in your function call, and to then specify the names of all optional arguments. If you don’t, someone reading your code might have to look up the definition of a function with unfamiliar arguments to understand what you’re doing.

# Vectors and data types

A vector is the most common and basic data type in R, and is pretty much the workhorse of R. A vector is composed by a series of values, which can be either numbers or characters. We can assign a series of values to a vector using the `c()` function. For example we can create a vector of animal weights and assign it to a new object `weight_g`:

```{r}
weight_g <- c(50, 60, 65, 82)
weight_g
```

A vector can also contain characters:

```{r}
animals <- c("mouse", "rat", "dog")
animals
```

The quotes around “mouse”, “rat”, etc. are essential here. Without the quotes R will assume objects have been created called `mouse`, `rat` and `dog`. As these objects don’t exist in R’s memory, there will be an error message.

There are many functions that allow you to inspect the content of a vector. `length()` tells you how many elements are in a particular vector:

```{r}
length(weight_g)
length(animals)
```

An important feature of a vector, is that all of the elements are the same type of data. The function `class()` indicates the class (the type of element) of an object:

```{r}
class(weight_g)
class(animals)

```

The function `str()` provides an overview of the structure of an object and its elements. It is a useful function when working with large and complex objects:

```{r}
str(weight_g)
str(animals)
```

You can use the `c()` function to add other elements to your vector:

```{r}
weight_g <- c(weight_g, 90) # add to the end of the vector
weight_g <- c(30, weight_g) # add to the beginning of the vector
weight_g
```

In the first line, we take the original vector `weight_g`, add the value 90 to the end of it, and save the result back into `weight_g`. Then we add the value `30` to the beginning, again saving the result back into  `weight_g`.

We can do this over and over again to grow a vector, or assemble a dataset. As we program, this may be useful to add results that we are collecting or calculating.

An atomic vector is the simplest R data type and is a linear vector of a single type. Above, we saw 2 of the 6 main atomic vector types that R uses: "character" and "numeric" (or "double"). These are the basic building blocks that all R objects are built from. The other 4 atomic vector types are:

    * "logical" for TRUE and FALSE (the boolean data type)
    * "integer" for integer numbers (e.g., 2L, the L indicates to R that it’s an integer)
    * "complex" to represent complex numbers with real and imaginary parts (e.g., 1 + 4i) and that’s all we’re going to say about them
    * "raw" for bitstreams that we won’t discuss further

You can check the type of your vector using the `typeof()` function and inputting your vector as the argument.

Vectors are one of the many data structures that R uses. Other important ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`), factors (`factor`) and arrays (`array`). The set of packages we are going to use a lot in this workshop, the `tidyverse` uses tibbles(`tibble`) which as sort of liked souped up data frames. Data frames are analogous to rectangular spreadsheets: they are representations of datasets in R where the rows correspond observations and the columns correspond to variables that describe the observations. 

<div class="box">

### EXERCISE

* We’ve seen that atomic vectors can be of type character, numeric (or double), integer, and logical. But what happens if we try to mix these types in a single vector?
 
* What will happen in each of these examples? (hint: use `class()` to check the data type of your objects):

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

* Why do you think it happens?

* How many values in `combined_logical` are "TRUE" (as a character) in the following example:

```{r}
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
combined_logical <- c(num_logical, char_logical)
```

You’ve probably noticed that objects of different types get converted into a single, shared type within a vector. In R, we call converting objects from one class into another class *coercion.* These conversions happen according to a hierarchy, whereby some types get preferentially coerced into other types. Can you draw a diagram that represents the hierarchy of how these data types are coerced?

</div>
 
 ```{r, echo=FALSE, eval= FALSE}
 # R implicitly converts them to all be the same type
 # Vectors can be of only one data type. R tries to convert (coerce) the content of this vector to find a “common denominator” that doesn’t lose any information.
 # Only one. There is no memory of past data types, and the coercion happens the first time the vector is evaluated. Therefore, the TRUE in num_logical gets converted into a 1 before it gets converted into "1" in combined_logical.
 # logical → numeric → character ← logical
```

# Subsetting vectors
If we want to extract one or several values from a vector, we must provide one or several indices in square brackets. For instance:

```{r}
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(3, 2)]
```

We can also repeat the indices to create an object with more elements than the original one:

```{r}
more_animals <- animals[c(1, 2, 3, 2, 1, 4)]
more_animals
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia, and R start counting at 1, because that’s what human beings typically do. Languages in the C family (including C++, Java, Perl, and Python) count from 0 because that’s simpler for computers to do.

# Conditional subsetting
Another common way of subsetting is by using a logical vector. `TRUE` will select the element with the same index, while `FALSE` will not:

```{r}
weight_g <- c(21, 34, 39, 54, 55)
weight_g[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

Typically, these logical vectors are not typed by hand, but are the output of other functions or logical tests. For instance, if you wanted to select only the values above 50:

```{r}
weight_g > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

You can combine multiple tests using & (both conditions are true, AND) or | (at least one of the conditions is true, OR):

```{r}
weight_g[weight_g < 30 | weight_g > 50]
weight_g[weight_g >= 30 & weight_g == 21]
```

Here, `<` stands for “less than”, `>` for “greater than”, `>=` for “greater than or equal to”, and `==` for “equal to”. The double equal sign `==` is a test for numerical equality between the left and right hand sides, and should not be confused with the single = sign, which performs variable assignment (similar to `<-`).

A common task is to search for certain strings in a vector. One could use the “or” operator | to test for equality to multiple values, but this can quickly become tedious. The function %in% allows you to test if any of the elements of a search vector are found:

```{r}
animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals %in% c("rat", "cat", "dog", "duck", "goat")
animals[animals %in% c("rat", "cat", "dog", "duck", "goat")]
```


<div class="box">
### EXERCISE
 
* Can you figure out why "four" > "five" returns TRUE?

</div>

```{r, eval=FALSE, echo = FALSE}
#When using “>” or “<” on strings, R compares their alphabetical order. Here “four” comes after “five”, and therefore is “greater than” it.
```

# Missing data
As R was designed to analyze datasets, it includes the concept of missing data (which is uncommon in other programming languages). Missing data are represented in vectors as `NA`.

When doing operations on numbers, most functions will return `NA` if the data you are working with include missing values. This feature makes it harder to overlook the cases where you are dealing with missing data. You can add the argument na.rm = TRUE to calculate the result while ignoring the missing values.

```{r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

If your data include missing values (and that's probably all of us!), you may want to become familiar with the functions `is.na()`,  `na.omit()`, and `complete.cases()`. See below for examples.

```{r}
## Extract those elements which are not missing values.
heights[!is.na(heights)]

## Returns the object with incomplete cases removed. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
na.omit(heights)

## Extract those elements which are complete cases. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
heights[complete.cases(heights)]

```

Recall that you can use the `typeof()` function to find the type of your atomic vector.


<div class="box">
### EXERCISE

1. Using this vector of heights in inches, create a new vector, `heights_no_na`, with the NAs removed.

```{r}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

2. Use the function median() to calculate the median of the heights vector.

3. Use R to figure out how many people in the set are taller than 67 inches.
</div>

```{r, eval = FALSE, echo = FALSE}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)

# 1.
heights_no_na <- heights[!is.na(heights)] 
# or
heights_no_na <- na.omit(heights)
# or
heights_no_na <- heights[complete.cases(heights)]

# 2.
median(heights, na.rm = TRUE)

# 3.
heights_above_67 <- heights_no_na[heights_no_na > 67]
length(heights_above_67)
```

# What are R packages
This section is from ["Modern Dive, section 2.3"](https://moderndive.com/2-getting-started.html#)

R packages extend the functionality of R by providing additional functions, data, and documentation. They are written by a world-wide community of R users and can be downloaded for free from the internet. A good analogy for R packages is they are like apps you can download onto a mobile phone

So R is like a new mobile phone: while it has a certain amount of features when you use it for the first time, it doesn’t have everything. R packages are like the apps you can download onto your phone from Apple’s App Store or Android’s Google Play.

Let’s continue this analogy by considering the Instagram app for editing and sharing pictures. Say you have purchased a new phone and you would like to share a recent photo you have taken on Instagram. You need to:
1. Install the app: Since your phone is new and does not include the Instagram app, you need to download the app from either the App Store or Google Play. You do this once and you’re set. You might do this again in the future any time there is an update to the app.   
2. Open the app: After you’ve installed Instagram, you need to open the app.   

Once Instagram is open on your phone, you can then proceed to share your photo with your friends and family. The process is very similar for using an R package. You need to:

1. *Install the package*: This is like installing an app on your phone. Most packages are not installed by default when you install R and RStudio. Thus if you want to use a package for the first time, you need to install it first. Once you’ve installed a package, you likely won’t install it again unless you want to update it to a newer version.   
2. *“Load” the package*: “Loading” a package is like opening an app on your phone. Packages are not “loaded” by default when you start RStudio on your computer; you need to “load” each package you want to use every time you start RStudio.

Let’s now show you how to perform these two steps for the `tidyverse` metapackage (package of packages).

<div class = "boxGrey">
There are two ways to install an R package. For example, to install the `tidyverse` package:

1. **Easy way**: In the Files pane of RStudio:
    a. Click on the “Packages” tab
    b. Click on “Install”
    c. Type the name of the package under “Packages (separate multiple with space or comma):” In this case, type `tidyverese`
    d. Click “Install”
    
2. **Slightly harder way**: An alternative way to install a package is by typing `install.packages("tidyverse")` in the Console pane of RStudio and hitting enter. Note you must include the quotation marks.    

Much like an app on your phone, you only have to install a package once. However, if you want to update an already installed package to a newer version, you need to re-install it by repeating the above steps.

</div>

### EXERCISE

Practice what you have just done and install the R package `here`.

</div>


### Package loading
Recall that after you’ve installed a package, you need to “load” it, in other words open it. We do this by using the `library()` command. For example, to load the `tidyverse` package, run the following code in the Console pane. What do we mean by “run the following code”? Either type or copy & paste the following code into the Console pane and then hit the enter key.

```{r}
library(tidyverse)
library(here)
```

If after running the above code, a blinking cursor returns next to the `>` “prompt” sign, it means you were successful and the `tidyverse` package is now loaded and ready to use. If however, you get a red “error message” that reads…

```
Error in library(tidyverse) : there is no package called ‘tidyverse’
```
it means that you didn’t successfully install it. In that case, go back to the previous subsection and trying install it again.


### Package use
One extremely common mistake new R users make when wanting to use particular packages is they forget to “load” them first by using the `library()` command we just saw. Remember: you have to load each package you want to use every time you start RStudio. If you don’t first “load” a package, but attempt to use one of its features, you’ll see an error message similar to:

```
Error: could not find function
```

R is telling you that you are trying to use a function in a package that has not yet been “loaded.” Almost all new users forget do this when starting out, and it is a little annoying to get used. to However, you’ll remember with practice.

# Attribution
This lesson was created by [Aleeza Gerstein](https://microstatslab.ca) at the University of Manitoba. It is sourced from material from [The Carpentries](https://carpentries.org/) [introductory lesson](https://datacarpentry.org/R-ecology-lesson/01-intro-to-r.html). Made available under the [Creative Commons Attribution license](https://creativecommons.org/licenses/by/4.0/). [License](https://datacarpentry.org/R-genomics/LICENSE.html).
